---
title: "Temporal and sequential analysis for learning analytics"
output: html_notebook
author: Quan Nguyen, Postdoctoral Fellow, School of Information, University of Michigan
---


# Day 1: Sequence analysis using TraMineR - EDA
# Day 2: Sequence analysis using TraMineR - Matching, clustering
# Day 3: Association rule mining using priori algorithm

## Loading libaries

```{r}
library(ggplot2)
library(data.table)
library(TraMineR)
library(dplyr)
library(tidyr)
```

```{r}
library(TraMineR)
data(mvad)
seqstatl(mvad[, 17:86])
mvad.alphabet <- c("employment", "FE", "HE", "joblessness", "school", 
    "training")
mvad.labels <- c("employment", "further education", "higher education", 
    "joblessness", "school", "training")
mvad.scodes <- c("EM", "FE", "HE", "JL", "SC", "TR")
mvad.seq <- seqdef(mvad, 17:86, alphabet = mvad.alphabet, states = mvad.scodes, 
    labels = mvad.labels, xtstep = 6)

```
```{r}
head(mvad)
```

```{r}
seqiplot(mvad.seq, with.legend = FALSE, title= "Index plot (10 first sequences")
```
```{r}
seqIplot(mvad.seq, sortv = "from.start", with.legend = FALSE)
```
```{r}
seqfplot(mvad.seq, withlegend = F, title = "Sequence frequency plot bar width proportional to the frequencies")
```
```{r}
 seqdplot(mvad.seq, withlegend = F,  title = "State distribution plot")
```


# Import log data

```{r}
lasi21_logdata <- read_csv("lasi21_logdata.csv")
lasi21_logdata$site_type <- ifelse(grepl("ssignment", lasi21_logdata$instancename), "assignment",lasi21_logdata$site_type )
lasi21_logdata$site_type <- ifelse(grepl("exam", lasi21_logdata$instancename), "assignment",lasi21_logdata$site_type )

lasi21_logdata <- as.data.table(lasi21_logdata[,c("id","date_time","spent_time","site_type","instancename")])
head(lasi21_logdata)
```

# Data pre-processing
```{r}
# convert ms to minutes
lasi21_logdata$spent_time_m <- round(lasi21_logdata$spent_time/60000, digits=1)

# create a flag for session break (aka where time spent > 30 minutes)
lasi21_logdata$session_flag = 0
lasi21_logdata$session_flag <- ifelse(lasi21_logdata$spent_time_m>30, 1, lasi21_logdata$session_flag)

# filter out all spent_time < 6s
lasi21_logdata2 <- lasi21_logdata[spent_time>=6000,]

# create session number 
lasi21_logdata2 <- lasi21_logdata2 %>% 
                        arrange(id,date_time,spent_time) %>% 
                        mutate(session_num = cumsum(session_flag))
# remove session break
lasi21_logdata2 <- lasi21_logdata2[session_flag==0,]


# for each learning session, calculate the cummulative time spent
lasi21_logdata2 <- lasi21_logdata2 %>% 
                        arrange(id,date_time,spent_time) %>% 
                        group_by(session_num) %>% 
                        mutate(spent_time_m_cum = cumsum(spent_time_m))

# create time unit as 1/10 of a minute (6s)
lasi21_logdata2$time_unit <- round(lasi21_logdata2$spent_time_m_cum*10,digits=0)


hist(lasi21_logdata2$spent_time_m_cum, main="Learning session length", xlab = "Minutes", breaks=200)
```

# Transform data into STS format
```{r}
# create a function to expand time unit for each learning session

f1 <- function(x1){
    x1 <- 1:max(x1)
    m1 <- max(c(length(x1)))
    length(x1) <- m1
    list(time_unit = x1)
}

# create a subset
lasi21_logdata3 <- lasi21_logdata2[,c("session_num","time_unit","site_type")]

# create an expanded df
test <- setDT(lasi21_logdata3)[, f1(time_unit), .(session_num)]

# merge with log data
test <- merge(test,lasi21_logdata3,by=c("session_num","time_unit"),all.x=TRUE)

# fill missing upward
test <- test %>% fill(site_type,.direction = "up")
```



```{r}
# Sequence length distribution
hist(lasi21_logdata3[,max(time_unit), by="session_num"]$V1, breaks=100, main="Length of learning session", xlab='Sequence length')
```

```{r}
# cummulative plot of seq length
plot(ecdf(lasi21_logdata3[,max(time_unit), by="session_num"]$V1), main ='Cummulative distribution of seq length')
```
# Define sequences

```{r}
# convert to STS format (wide format)
log_sts <- spread(test, time_unit, site_type)

# define sequences columns
log_sts.seq <- seqdef(log_sts,2:1200)
```

# EDA

```{r}
# plot the first 10 sequences with length=100
layout(matrix(c(1,1,2,1,1,2,3,4,4), nrow = 1, ncol = 3, byrow = TRUE))
seqiplot(log_sts.seq[611:621,1:100], withlegend = F, title = "Index plot (10 first sequences)")
seqlegend(log_sts.seq)
```


```{r}
# State distribution plot
layout(matrix(c(1,1,2,1,1,2,3,4,4), nrow = 1, ncol = 3, byrow = TRUE))
seqdplot(log_sts.seq[,1:200], title = "State distribution plot", withlegend = FALSE)
seqlegend(log_sts.seq)
```

```{r}
# Sequence frequency plot

layout(matrix(c(1,1,2,1,1,2,3,4,4), nrow = 1, ncol = 3, byrow = TRUE))
seqfplot(log_sts.seq[,1:200], title = "Sequence frequency plot", withlegend = FALSE, pbarw = TRUE)
seqlegend(log_sts.seq)
```
```{r}
# Stability within sequences
# Shannon's entropy as a measure of the diversity of states observed at the considered time point
# It equals 0 when all cases are in the same state  (it is thus easy to predict in which state an individual is)
# It is maximum when the cases are equally distributed between the states
seqHtplot(log_sts.seq[,1:200], title = "Entropy index")
```

```{r}
# Turbulence
Turbulence <- seqST(log_sts.seq[,1:200])
summary(Turbulence)
hist(Turbulence, col = "cyan", main = "Sequence turbulence",breaks=50)

```

```{r}
# Transitions of events

# define seq transitions
log_sts.seqe <- seqecreate(log_sts.seq[,1:200])

# find frequent subsequences
fsubseq <- seqefsub(log_sts.seqe, pMinSupport = 0.05)

# plot 15 most frequent subsquences
plot(fsubseq[1:15], col = "cyan", main="Top 15 frequent subsequences")

```





Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

